---
title: "Week 1 AUG 25-AUG 31 Lab 1 Introduction to R"
author: "Lysa Vanible"
date: "`r format(Sys.Date(), format='%B %d, %Y')`"
output:
 pdf_document:
  number_sections: true
---

# Introduction 
Read carefully through this introduction before you start work on this lab. 

1)Do not delete anything in this lab document, unless it is written by you or the assignment specifically requests you to do that. If you delete anything that was not written by you, you could end up removing important instructions or code from the document. If you do that, you may not be able to complete the lab successfully.

2). This first lab introduces you to R and Rstudio and rmarkdown. You are going to read and work through multiple tasks. Some of them require you to run R code provided by the instructor but some require you to enter code in the document and run it.  Whenever you are asked to write your own code, use the example codes that I provide to develop your own. Make sure to write all code INSIDE THE GREY BOXES/CHUNKS. DO NOT WRITE ANY CODES OUTSIDE THE CHUNKS. BE AWARE THAT R IS VERY PICKY AND SO YOU MUST FOLLOW CAPITALIZATION, SPELLING AND PUNCTUATION EXACTLY.

3). There are two ways to run R code in this document.
  a). One way is to highlight the code you want to run by clicking the row number next to the code. Once highlighted, click "Run" at the right top of this window and select "Run Selected Line(s)". The output or results will usually be produced beneath the chunk. Sometimes however the results will be stored in an object and will not be produced on screen. 
  b). A second way is to click on the tiny green arrow at the top right of the  CHUNK. Approach (a) will run only the highlighted code. Approach ( b) will execute all the code in the chunk.
 c). Consequently, once the code is written, if the assignment requests that you run the code, highlight it, click on "Run" and select"Run Selected Line(s)" or click on the green arrow at the top right corner of the chunk. It is VERY VERY important that you follow instructions precisely. 

4). Once you are done with all the tasks required in the lab and get to the conclusion, you will knit the assignment into a pdf/html document by clicking on the "Knit" command at the top of this window. Instructions are provided in the conclusion to this assignment.

5). Before you go on - please write your name in the YAML header at the top of this page - where it says author. Replace "ENTER YOUR NAME" with your name. Then skip to "Part A Rudimentary R Moves" and continue reading and following the instructions.

```{r setup, cache = F}
#This command tells R to continue Knitting the document even if there are code errors

knitr::opts_chunk$set(error = TRUE)


```

PART A RUDIMENTARY MOVES  
Let us use an example to clarify what R can do. To calculate 1+1 in R you enter and run the expression 1+1 in the chunk below. For now, the 1+1 expression has been written for you in the chunk. Run the expression to compute it. Highlight the code 1+1 by clicking on the row number next to it and then clicking "Run" at the top right of this window and selecting "Run Selected Line(s)".  R will run the expression and if there are no errors, it will spit out the answer as 2 below the chunk.  You can also click on the tiny green arrow at the top right of the chunk to run the code.

```{r}



1+1



```

R can also store data in variables. An example of this is provided below. The variable "x" is created and the assignment symbol "<-" is used to tell R to store 50 in it. Go ahead and run the code below to store 50 in x. To run, highlight the code by clicking on the row number next to it and then clicking "Run" and selecting "Run Selected Line(s)".  

```{r}



x<-50
x




```

You do not see any results because R stored the number 50 in x. To see that R did that, run the code below to write out the contents of x. Highlight the row number next to x and click Run. R will spit out the number stored in x below the chunk.

```{r}



x



```

Now it is your turn to do this. Follow the example above to write code in the chunk below that will store 300 in y. Then highlight the code and use the "Run" command to execute it. Remember that R is case sensitive so y is not the same as Y to R.
```{r}
#write code to store 300 in y below this line
y<-300
y





```

R will remember the value of x from before. So we can make computations using the variable x. The expression below tells R to multiply the contents of x by 4. The asterisk symbol means multiply. Go ahead and calculate the expression below by highlighting the code and using "Run" or the green arrow to execute. 

```{r}


x*4


```

R will also remember the value you assigned to y earlier and can use it in computations. Find out what x minus y is by entering the appropriate expression in the grey box below and calculating it in R?

```{r}
#enter the expression x-y below this line and run it

x-y






```

PART B WORKING WITH DATASET "state.x77    
Part A introduced issuing simple commands in R. Part B continues this introduction by getting you to work with a dataset called 'state.x77'. The dataset contains multiple pieces of information about all the states in the U.S. It came with R software when it was installed on the server. First let us look at the contents of state.x77. Run the code below to see the contents. R will spit out the contents below the chunk. You wll see that the data set comprises the names of the states and information for them about Population(1975 population in 1000s), Income(1974 per capita income), Illiteracy(1970 percent of population illiterate), Life Exp(1969-71 life expectancy at birth in years), Murder(murders and non-negligent manslaughters per 100,000 people in 1976), HS Grad(1970 percent high school graduates), Frost(mean number of days with minimum temperature below freezing (1931–1960) in capital or large city),  Area(land area of state in square miles). We are going to use these data to illustrate more of the things R can do.
```{r}



state.x77



```

PART C EXAMINING STATE POPULATION SIZES   
Now we want to perform some computations on state.x77. To get R to perform computations on state.x77 we will first use the code below to prepare the data,including renaming it "states". Run the code below to prepare the data.
```{r}




states<-data.frame(state.x77)




```

Now we will use the data "states" to examine the state population sizes. To tell R to use a particular variable, we first tell R the dataset with the variable, then use a $ symbol to declare the variable. Lets use R to inspect population sizes among U.S. states. For instance, let us see if R can tell which state is the largest population and which has the least population. First the state with the largest population. Run the code below to see that California had the largest population in 1975. A population size of 21,198,000
```{r}




states[which.max(states$Population),]




```
Now the state with the least population? Run the code below to find that Alaska had the smallest population size in 1975. 
```{r}




states[which.min(states$Population),]





```
How about we compute the population size of the average state by running the R code below?
```{r}




mean(states$Population)




```
Doing that tells you that the population size of the average state is 4,246,420. Obviously, population varies considerably by state. While the average state has a population size of 4,246,420 people. This varies from a minimum of 365,000 people in Alaska to 21,198,000 in California.

PART D WORKING WITH PER CAPITA INCOME  
Now let's examine per capita income, which is the average income per person, for the various states. Review the way I handled the tasks dealing with population size in part C and then find the states with the highest and lowest per capita income as well as the mean state per capita income for 1970. To find the state with the highest per capita income, follow my examples above and enter the appropriate code in the chunk below and run it. The relevant variable in the dataset is "Income" without the quotes. R is very picky so make sure you get the capitalization, spelling and punctuation exactly right to avoid errors. 
```{r}
#enter your code below this line and run it to find the state with the highest income per capita



states[which.max(states$Income),]



```

To find the state with the lowest per capita income, follow my example above and enter and run the appropriate code below. 
```{r}
#enter your code below this line and run it to find the state with the lowest income per capita


states[which.min(states$Income),]





```
 
How about the per capita income of the average state?
 
```{r}
#enter your code below this and run it to find the per capita income of the average state

mean(states$Income)









```
 
PART E COMPUTING POPULATION DENSITY  
Now we will use the data "states" to compute the population density of each state. Population density tells us how crowded a state is. It is expressed as the number of people per square mile. The higher the density the more crowded a state is. To calculate population density for a state, we divide the state's population by its square area. The population figures are in 1000s so we will need to multiply the computation by 1000s to get the right population density. We will also create a new variable called Population_Density to store the results of the density calculations. We use the underscore to connect Population and Density because R has problems with names that have spaces inside them. Calling the new variable states$Population_Density tells R to add population density as a new column in the states dataset. 

Run the following code to compute population density for all the states and add it to the states dataset.
```{r}



states$Population_Density <- states$Population*1000/states$Area




```
Now let us inspect 'states' to learn more about population density among U.S. states. For instance, let us see if we can tell which state is the most crowded and which is the least crowded and what their densities are. Enter the appropriate code below to find the state with the highest population density. The results will appear below the code chunk. Remember that the variable for population density is "Population_Density" without the quotes.
```{r}
#enter your code below this and run it to find the state with the highest population density
 
 states$Population_Density 
 
 
 
 
 
 
```
How about the least densely populated state? Enter and run the appropriate code to find that below? 
```{r}
#enter your code below this and run it to find the state with the lowest population density 
 
states[which.min(states$Population_Density),]
 

 
 
 
 
 
```
What is the population density for average state? Enter and run the appropriate R code below.
```{r}
#enter your code below this and run it to find the population density for the average state
 
 mean(states$Population_Density)
 

 
 
 
 
```
PART F. CONCLUSION  
Great work!!!! You have a few more things to do but you are almost at the end of your first lab. 

1). You will now prepare to knit the lab. Click on the "Knit" command at the top of this window to turn this document with your code and all the results into a pdf document. It will take a few seconds. When the knit is complete, Rstudio will open the pdf document. Inspect it very carefully to make sure that you have not missed any part of the assignment. If you have missed any tasks please go back and fix them and knit the document again. Once the pdf document is knitted and you are sure you have completed all the coding tasks in it prepare to submit it to the week 1 folder of SOC339's blackboard site. Before you can do that there are a few more things for you to do. 

2). First save your rmarkdown file (this file you have been working on in Rstudio) by clicking on the FILE command in the Rstudio menu and then on SAVE. You will see the rmarkdown document in the FILES folder of the window to the bottom right of your screen. It will be named Lab 1 Introduction to R.Rmd. Leave it there. 

3). You will also see the pdf document you knitted in the FILES window. It will be named Lab 1 Introduction to R.pdf. This document has to be uploaded into blackboard. To do that click on the little box to the left of Lab 1 Introduction to R.pdf. Then click on the MORE tab near the top of the FILES window.  Then on the EXPORT and then when prompted by RSTUDIO click on DOWNLOAD. RSTUDIO will save Lab 1 Introduction to R.pdf to the DOWNLOAD folder of your computer.

4). Go to the blackboard Week 1 folder and follow the instructions to complete lab 1 and upload the Lab 1 Introduction to R.pdf. 